Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-06-26 draft.tex
proofread first half

We integrate an \emph{overview visualization}, clustering of features and object shape for data-driven subgroup definition and statistical key figures for measuring the association between features.

@@

\maketitle
%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
Observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examination results that may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis are employed to check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of people, randomly selected to avoid any bias.
%
The subjects are selected without focus on a certain disease.
%
A large cohort size is essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
% 
The concurrent analysis of image data and non-spatial epidemiological factors requires techniques that reach beyond standard statistical methods.
%
For instance, segmentation of the image data is required for an analysis of anatomical structure and of possible correlations between this structure and epidemiological factors.
%
Semi-automatic segmentation techniques are promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
% Segmentation of these data enables an anatomical structure analysis and an analysis of possible correlations with other features.
% %
% Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%

@@

\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/workflow_comparison}
 \caption{
 (a) The standard epidemiology workflow consists of four steps.
 %
 (b) \emph{IVA} tools complement parts of this workflow instead of replacing them.
 %
 The combination of statistical and interactive-driven analysis shows promising potential to unveil information in the data.
 %
 %IVA complements this workflow and is used to elaborate the feature selection.
 %
 We call the iterative red highlighted part \emph{IVA Loop}, described in detail in Figure~\ref{fig:InteractionLoop}.
 %(a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.
 %
 }
  \label{fig:WorkflowComparison}
\end{figure}
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew et al. \cite{Thew2009}, the workflow can be characterized as follows:
%
\begin{enumerate}\itemsep0.1em
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
%
	\item Confounding features are identified and taken into account (for example using stratification).
%
	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.
%
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach. %although it does not consider image data.
%
We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.

\emph{Reproducibility} of results is an epidemiological key requirement.
%
It is difficult to achieve, since many physicians are involved when thousands of test persons are examined and interviewed.
%
Thus, both intra- and inter-observer variability needs to be low for all aspects of a cohort study examination.
%
Longitudinal studies require the acquired features to be comparable for evaluation.
%
%If the data acquisition process changes, an information bias is introduced to the data, hampering inference in and between acquisition cycles.
%
%
Grouping subjects using epidemiological features is essential in cohort studies to allow per-group risk determination.
%
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20 year steps) when investigating its influence.
%
These groups strongly depend on the condition of interest and therefore there is no standard for their categorization.
% --- NEW PARAGRAPH

@@

Our work is closest to that of Steenwijk \cite{Steenwijk2010}, Turkey \cite{Turkay2013}, and Angelelli \cite{Angelelli2014} and colleagues, who employ multiple coordinated view systems for the analysis of cohort study.
%
%They employ multiple coordinated view systems for the analysis of cohort study data and demonstrate their approaches based on an aging study \cite{Turkay2013, Angelelli2014} and a cohort of patients suspected of having a special type of rheumatic disease \cite{Steenwijk2010}.

@@

Gresh et al. \cite{Gresh2000} proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views.
%
Blaas et al. \cite{Blaas2007} presented a similar system, which also analyzed medical image data and features derived from them using views from the feature and physical space.
%
Both works are restricted to the analysis of one case at a time and to non-image data with a unique spatial reference, e.g., voltage simulated across the heart muscle.
%
In epidemiology, multiple cases must be concurrently investigated and non-image data often lacks a spatial reference, such as gender and age.
%
% \com{\textbf{Reviewer 1}: The related work section includes merely a set of related tools. I would have liked to see a more comprehensive section that discusses the previous related work in more detail and describes why they are important and relevant to this work and how your work differs or improves upon theirs.\\
% %
% Quality: The work contains many references. But there are potentially some missing; e.g, the framework talks about utilizing multiple coordinated views (although the details of what is actually linked is vague) and therefore cited papers such as Baldonado's work in AVI 2000, or Roberts 2007 on Coordinated multiple views, or Weaver's Cross-filtered Views for multidimensional visual analysis 2010 should be cited. The paper really proposes a framework; however it was unclear how this was implemented into the presented system. Therefore this should be improved.\\\\
% %
% \textbf{Reviewer 2}
% - I would suggest to also cite the original Generalized Pairs Plot and
% not only the adaptation as GPLOMS [21]: J. W. Emerson and W. A. Green.
% gpairs: The Generalized Pairs Plot, 2012. R package version 1.1.\\
% %
% - A related process model for Visual Analytics that focuses on hypotheses
% and models: T. Lammarsch, W. Aigner, A. Bertone, S. Miksch, and A. Rind,
% “Towards a Concept how the Structure of Time can Support the Visual
% Analytics Process,” in Proceedings of International Workshop on Visual
% Analytics (EuroVA 2011) in conjunction with EuroVis 2011, Goslar,
% Germany, 2011, pp. 9–12.\\
% %
% - VISITORS is related work in visualization of medical variables of
% cohorts: D. Klimov, Y. Shahar, and M. Taieb-Maimon, “Intelligent
% visualization and exploration of time-oriented data of multiple
% patients,” Artificial Intelligence in Medicine, vol. 49, no. 1, pp.
% 11–31, May 2010.\\
% %
% - another, forthcoming related paper: P. Angelelli, S. Oeltze, C. Turkay,
% J. Haasz, E. Hodneland, A. Lundervold, A. Lundervold, B. Preim, and H.
% Hauser, “Interactive Visual Analysis of Heterogeneous Cohort Study
% Data,” IEEE Computer Graphics and Applications, vol. Early Access
% Online, 2014.\\\\
% %
% \textbf{Reviewer 3}
% What the prior work section needs at the end is a paragraph stating how their work differs from the work they summarise i.e. it uses more non-image data than Turkay and uses different visualisations, and more importantly differentiation from Steenwijk et al’s very similar work (saying they use a different meaning for the word “cohort” does not seem justification enough, given the large overlap), only then can we be sure there is a novel contribution here.}

@@

Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.

@@

While their goal to identify possible risk factors using socio-economic and health data similar to ours, they focus on iteratively refining defined hypothesis and on geographical data.

@@

While the works shows, how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.

>>>>

We integrate an overview visualization, clustering of features and object shape for data-driven subgroup definition and statistical key figures for measuring the association between features.

@@

\maketitle
%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
Observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examination results that may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis are employed to check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of people, randomly selected to avoid any bias.
%
The subjects are selected without focus on a certain disease.
%
A large cohort size is essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
% TODO - Khonya Paper Venue
% TODO - Introduction Highlight Overview visualization?
% TODO - Klemm et al. - Klingt ein bisschen shizophren
Cohort studies often include medical image data.
% 
The concurrent analysis of image data and non-spatial epidemiological factors requires techniques that reach beyond standard statistical methods.
%
For instance, segmentation of the image data is required for an analysis of anatomical structure \com{and of possible} correlations between this structure and epidemiological factors.
%
Semi-automatic segmentation techniques are promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
% Segmentation of these data enables an anatomical structure analysis and an analysis of possible correlations with other features.
% %
% Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%

@@

\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/workflow_comparison}
 \caption{
 (a) The standard epidemiology workflow consists of four steps.
 %
 (b) \emph{IVA} tools complement parts of this workflow instead of replacing them.
 %
 The combination of statistical and interactive analysis shows promising potential to unveil information in the data.
 %
 %IVA complements this workflow and is used to elaborate the feature selection.
 %
 We call the iterative red highlighted part \emph{IVA Loop}, described in detail in Figure~\ref{fig:InteractionLoop}.
 %(a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.
 %
 }
  \label{fig:WorkflowComparison}
\end{figure}
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics, and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew et al. \cite{Thew2009}, the workflow can be characterized as follows:
%
\begin{enumerate}\itemsep0.1em
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
%
	\item Confounding features are identified and taken into account (for example using stratification).
%
	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.
%
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach. %although it does not consider image data.
%
We focus, on the potential of image data and attempt to support \emph{hypothesis generation}.

\emph{Reproducibility} of results is an epidemiological key requirement.
%
It is difficult to achieve, since many physicians are involved when thousands of subjects are examined and interviewed.
%
Thus, both intra- and inter-observer variability needs to be low for all aspects of a cohort study examination.
%
Longitudinal studies require the acquired features to be comparable for evaluation.
%
%If the data acquisition process changes, an information bias is introduced to the data, hampering inference in and between acquisition cycles.
%
%
Grouping subjects using epidemiological features is essential in cohort studies to allow per-group risk determination.
%
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20 year steps) when investigating its influence.
%
These groups strongly depend on the condition of interest and therefore there is no standard for their categorization.
% --- NEW PARAGRAPH

@@

Our work is closest to that of Steenwijk \cite{Steenwijk2010}, Turkey \cite{Turkay2013}, and Angelelli \cite{Angelelli2014} and colleagues, who employ multiple coordinated view systems for the analysis of cohort study data.
%
%They employ multiple coordinated view systems for the analysis of cohort study data and demonstrate their approaches based on an aging study \cite{Turkay2013, Angelelli2014} and a cohort of patients suspected of having a special type of rheumatic disease \cite{Steenwijk2010}.

@@

Gresh et al. \cite{Gresh2000} proposed \texttt{WEAVE}, one of the first systems which concurrently analyzed medical image and non-image data using linked views.
%
Blaas et al. \cite{Blaas2007} presented a similar system, which also analyzed medical image data and features derived from them using views from the feature and physical space.
%
Both works are restricted to the analysis of one case at a time and to non-image data with a unique spatial reference, e.g., voltage simulated across the heart muscle.
%
In epidemiology, multiple cases must be concurrently investigated and non-image data often lacks a spatial reference, such as gender or age.
%
% \com{\textbf{Reviewer 1}: The related work section includes merely a set of related tools. I would have liked to see a more comprehensive section that discusses the previous related work in more detail and describes why they are important and relevant to this work and how your work differs or improves upon theirs.\\
% %
% Quality: The work contains many references. But there are potentially some missing; e.g, the framework talks about utilizing multiple coordinated views (although the details of what is actually linked is vague) and therefore cited papers such as Baldonado's work in AVI 2000, or Roberts 2007 on Coordinated multiple views, or Weaver's Cross-filtered Views for multidimensional visual analysis 2010 should be cited. The paper really proposes a framework; however it was unclear how this was implemented into the presented system. Therefore this should be improved.\\\\
% %
% \textbf{Reviewer 2}
% - I would suggest to also cite the original Generalized Pairs Plot and
% not only the adaptation as GPLOMS [21]: J. W. Emerson and W. A. Green.
% gpairs: The Generalized Pairs Plot, 2012. R package version 1.1.\\
% %
% - A related process model for Visual Analytics that focuses on hypotheses
% and models: T. Lammarsch, W. Aigner, A. Bertone, S. Miksch, and A. Rind,
% “Towards a Concept how the Structure of Time can Support the Visual
% Analytics Process,” in Proceedings of International Workshop on Visual
% Analytics (EuroVA 2011) in conjunction with EuroVis 2011, Goslar,
% Germany, 2011, pp. 9–12.\\
% %
% - VISITORS is related work in visualization of medical variables of
% cohorts: D. Klimov, Y. Shahar, and M. Taieb-Maimon, “Intelligent
% visualization and exploration of time-oriented data of multiple
% patients,” Artificial Intelligence in Medicine, vol. 49, no. 1, pp.
% 11–31, May 2010.\\
% %
% - another, forthcoming related paper: P. Angelelli, S. Oeltze, C. Turkay,
% J. Haasz, E. Hodneland, A. Lundervold, A. Lundervold, B. Preim, and H.
% Hauser, “Interactive Visual Analysis of Heterogeneous Cohort Study
% Data,” IEEE Computer Graphics and Applications, vol. Early Access
% Online, 2014.\\\\
% %
% \textbf{Reviewer 3}
% What the prior work section needs at the end is a paragraph stating how their work differs from the work they summarise i.e. it uses more non-image data than Turkay and uses different visualisations, and more importantly differentiation from Steenwijk et al’s very similar work (saying they use a different meaning for the word “cohort” does not seem justification enough, given the large overlap), only then can we be sure there is a novel contribution here.}

@@

Their definition of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.

@@

While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypothesis and on geographical data.

@@

While the work shows, how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
